---
title: "Reproducible Research: Peer Assessment 2"
output: 
  html_document:
    keep_md: true
---
## Synopsis  
Need to put summary here


## Load libraries
```{r libraries}
suppressPackageStartupMessages({
  require(data.table)
  require(dplyr)
  require(lubridate)
  require(ggplot2)
  require(grid)
  require(gridExtra)
  })
```


## Data load
```{r read_data, cache=TRUE}
raw_data <- read.csv('repdata%2Fdata%2FStormData.csv.bz2', stringsAsFactors = FALSE)
```

## Data Processing

The storm database has been built up over a number of years.   
The event type (EVTYPE) field is free text and contains entries  
that are very similar and some spelling errors. Entries that are  
similar have been grouped together (eg. everything containing  
'hurricane').  

```{r process_data}

fix_EVTYPE <- function(x, grep_string, replace_string) {
  idx <- grepl(grep_string, x$EVTYPE, ignore.case = TRUE )
  res <- x
  if( sum(idx) > 0 ) {
    cat('Found', paste0(unique(x[idx,'EVTYPE']), collapse=', '), '\n')
    cat('Replacing with', replace_string, '\n' )
    res[idx,'EVTYPE'] <- replace_string
  }
  return(res)
}

clean_data <- as.data.table(raw_data) %>%
                mutate(YEAR = year(mdy_hms(BGN_DATE))) %>%
                mutate(PROPDMGVAL = PROPDMG * ifelse( PROPDMGEXP == 'K', 1e-3,
                                              ifelse( PROPDMGEXP == 'M', 1,
                                              ifelse( PROPDMGEXP == 'B', 1e3, 0))))

clean_data <- fix_EVTYPE(clean_data, 'hurricane', 'HURRICANE')
clean_data <- fix_EVTYPE(clean_data, 'TSTM|thunderstorm', 'THUNDERSTORM')
clean_data <- fix_EVTYPE(clean_data, 'heat', 'HEAT')
clean_data <- fix_EVTYPE(clean_data, 'flood', 'FLOOD')
clean_data <- fix_EVTYPE(clean_data, 'cold|winter', 'COLD')
clean_data <- fix_EVTYPE(clean_data, 'rip', 'RIP')
```

## Selection of time period

Using recent data has 3 advantages  
1. Database is more complete  
2. Reflective of recent climatic conditions  
3. Reflective of CURRENT infrastructure and support  

The last one is the most important - we don't want to have a response to  
damage/fatalities that have previously occurred where infrastructure and/or  
support has already changed because of this.  

```{r analyse_time_series}
time_data <- group_by(clean_data, YEAR) %>%
               summarise(total_death  = sum(FATALITIES),
                         total_damage = sum(PROPDMGVAL)
                        )

death_plot <- qplot(YEAR,
                    total_death,
                    data = time_data,
                    ylab = 'total fatalities',
                    geom = c('point', 'line')
                   )      

damage_plot <- qplot(YEAR,
                     total_damage,
                     data = time_data,
                     ylab = 'total damage (millions)',
                     geom = c('point', 'line')
                    )

grid.arrange( death_plot, damage_plot, ncol=1)

recent_data <- filter( clean_data, YEAR >= 1993 )
```

## Results

HEAT causes most deaths on average over the considered time period.  
In addition, this type has highest number of deaths in a single year.  
FLOOD has next most deaths on average, although TORNADO has second highest number  
of deaths in a single year.  

```{r results}
# Cannot use mean function within summarise because this will group by event types
# and take the average over the series of individual events

yearly_data <- group_by(recent_data, YEAR, EVTYPE)        %>%
                 summarise( yearly_death = sum(FATALITIES))

total_data <- group_by(yearly_data, EVTYPE) %>%
                 summarise(total_death = sum(yearly_death),
                           avg_death   = mean(yearly_death))

death_data <- arrange(total_data, desc(total_death))

idx <- yearly_data$EVTYPE %in% death_data$EVTYPE[1:5]

qplot(EVTYPE,
      yearly_death,
      data = yearly_data[idx,],
      geom = 'boxplot',
      xlab = 'Event Type',
      ylab = 'Yearly Fatalities',
      main = 'Top 5 Event Types for fatalities since 1993'
      )

cat('NOTE: Boxplot removes outliers when determining average\n\n')

cat('Table of Average Yearly Fatalities for Top 5 Event Types\n')

print(death_data[1:5,])

#----------------------------------

yearly_data <- group_by(recent_data, YEAR, EVTYPE)        %>%
                 summarise( yearly_damage = sum(PROPDMGVAL))

total_data <- group_by(yearly_data, EVTYPE) %>%
                 summarise(total_damage = sum(yearly_damage),
                           avg_damage   = mean(yearly_damage))

damage_data <- arrange(total_data, desc(total_damage))

idx <- yearly_data$EVTYPE %in% damage_data$EVTYPE[1:5]

xxxxx
qplot(EVTYPE,
      total,
      data = damage_data,
      geom = 'boxplot',
      xlab = 'Event Type',
      ylab = 'Yearly Damage (millions)',
      main = 'Top 5 Event Types for property damage since 1993'
      )

cat('NOTE: Boxplot removes outliers when determining average\n\n')

cat('Table of Average Yearly Property Damage (millions) for Top 5 Event Types\n')
print(avg_damage_by_type[1:5,])

```
