---
title: "Since 1993, HEAT has caused most fatalities, while FLOOD has caused most property damage"
output: 
  html_document:
    keep_md: true
---
## Synopsis  
Analysis of the U.S. National Oceanic and Atmospheric Administation (NOAA) storm database shows that in the period since 1993, HEAT events have caused the most fatalities while FLOOD events have caused the most property damage. The period since 1993 has been chosen because that data is both more complete and more reflective of current infrastructure. The impact of major storms and weather events is highly variable and planning should consider both the peak yearly fatalities and damage as well as the total fatalities and damage. This investigation has not considered correlation between event types such as flooding directly caused by otherwise labelled storm events. For this analysis, event types have been combined however further consolidation and categorisation of event types in the database would be useful.

## Load libraries
```{r libraries}
suppressPackageStartupMessages({
  require(data.table)
  require(dplyr)
  require(lubridate)
  require(ggplot2)
  require(grid)
  require(gridExtra)
  })
```


## Data Processing
Data is loaded from the compressed database extraction.  

Only the 5 columns that are used are extracted.  

Event type (EVTYPE) entries that are similar have been grouped together (eg. everything containing 'hurricane'). The EVTYPE field is free text and appears to have been built up over the years. It contains entries that are almost the same and some abbreviations (eg. TSTM for Thunderstorm) as well as obvious spelling errors.

```{r load_data, cache=TRUE}
raw_data <- read.csv('repdata%2Fdata%2FStormData.csv.bz2',
                     colClasses = c('NULL',
                                    'character',     # BGN_DATE
                                    rep('NULL',5),
                                    'character',     # EVTYPE
                                    rep('NULL', 14),
                                    'numeric',     # FATALITIES
                                    'NULL',
                                    'numeric',       # PROPDMG
                                    'character',     # PROPDMGEXP
                                    rep('NULL', 11)
                                   ),
                     stringsAsFactors = FALSE)
```


```{r process_data}

fix_EVTYPE <- function(x, grep_string, replace_string) {
  idx <- grepl(grep_string, x$EVTYPE, ignore.case = TRUE )
  res <- x
  if( sum(idx) > 0 ) {
    cat('Found', paste0(unique(x[idx,'EVTYPE']), collapse=', '), '\n')
    cat('Replacing with', replace_string, '\n' )
    res[idx,'EVTYPE'] <- replace_string
  }
  return(res)
}

clean_data <- as.data.table(raw_data) %>%
                mutate(YEAR = year(mdy_hms(BGN_DATE))) %>%
                mutate(PROPDMGVAL = PROPDMG * ifelse( PROPDMGEXP == 'K', 1e-3,
                                              ifelse( PROPDMGEXP == 'M', 1,
                                              ifelse( PROPDMGEXP == 'B', 1e3, 0))))

clean_data <- fix_EVTYPE(clean_data, 'hurricane', 'HURRICANE')
clean_data <- fix_EVTYPE(clean_data, 'TSTM|thunderstorm', 'THUNDERSTORM')
clean_data <- fix_EVTYPE(clean_data, 'heat', 'HEAT')
clean_data <- fix_EVTYPE(clean_data, 'flood', 'FLOOD')
clean_data <- fix_EVTYPE(clean_data, 'cold|winter', 'COLD')
clean_data <- fix_EVTYPE(clean_data, 'rip', 'RIP')
```

## Selection of time period

Using recent data has 3 advantages  
1. Database is more complete  
2. Reflective of recent climatic conditions  
3. Reflective of CURRENT infrastructure and support  

The last one is the most important - we don't want to have a response to  
damage/fatalities that have previously occurred where infrastructure and/or  
support has already changed because of this.  

```{r analyse_time_series}
time_data <- group_by(clean_data, YEAR) %>%
               summarise(total_death  = sum(FATALITIES),
                         total_damage = sum(PROPDMGVAL)
                        )

death_plot <- qplot(YEAR,
                    total_death,
                    data = time_data,
                    ylab = 'fatalities',
                    geom = c('point', 'line')
                   )      

damage_plot <- qplot(YEAR,
                     total_damage,
                     data = time_data,
                     ylab = 'property damage (millions)',
                     geom = c('point', 'line')
                    )

grid.arrange( death_plot,
              damage_plot,
              ncol = 1,
              main = 'Time Series analysis of NOAA Storm Database'
            )

recent_data <- filter( clean_data, YEAR >= 1993 )
```
Time analysis of fatalities and property damage indicates that the number
of database entries has increased over recent years. As a result, only data
from 1993 inclusive has been used.

## Results

HEAT causes most deaths on average over the considered time period.  
In addition, this type has highest number of deaths in a single year.  
FLOOD has next most deaths on average, although TORNADO has second highest number  
of deaths in a single year.  

```{r death_results}

yearly_data <- group_by(recent_data, YEAR, EVTYPE)           %>%
                 summarise( yearly_death = sum(FATALITIES))

total_data <- group_by(yearly_data, EVTYPE)                  %>%
                 summarise(total_death = sum(yearly_death),
                           avg_death   = mean(yearly_death)) %>%
                 arrange(desc(total_death))

idx <- yearly_data$EVTYPE %in% total_data$EVTYPE[1:5]

qplot(EVTYPE,
      yearly_death,
      data = yearly_data[idx,],
      geom = 'boxplot',
      xlab = 'Event Type',
      ylab = 'Yearly Fatalities',
      main = 'Top 5 Event Types for fatalities since 1993'
      )

cat('NOTE: Boxplot removes outliers when determining average\n\n')

cat('Table of Average Yearly Fatalities for Top 5 Event Types\n')

print(total_data[1:5,])
```

Add information about damage results


```{r damage_results}

yearly_data <- group_by(recent_data, YEAR, EVTYPE)             %>%
                 summarise( yearly_damage = sum(PROPDMGVAL))

total_data <- group_by(yearly_data, EVTYPE)                    %>%
                 summarise(total_damage = sum(yearly_damage),
                           avg_damage   = mean(yearly_damage)) %>%
                           arrange(desc(total_damage))

idx <- yearly_data$EVTYPE %in% total_data$EVTYPE[1:5]

qplot(EVTYPE,
      yearly_damage,
      data = yearly_data[idx,],
      geom = 'boxplot',
      xlab = 'Event Type',
      ylab = 'Yearly Property Damage (millions)',
      main = 'Top 5 Event Types for property damage since 1993'
      )

cat('NOTE: Boxplot removes outliers when determining average\n\n')

cat('Table of Average Yearly Property Damage (millions) for Top 5 Event Types\n')
print(total_data[1:5,])

```
